---
title: "Possible background artifact on arrays"
author: "João Baiochi, baiochi@usp.br"
date: "August 29, 2015"
output: 
  html_document: 
    theme: flatly
    toc: true
    highlight: tango
---

```{r Pre Processing, echo=FALSE, warning=FALSE, results='hide'}
library(limma)
library(knitr)
source("/Users/Baiochi/Desktop/Agilent-Bioc/Scripts/Functions.R")
targets <- read.table(file="/Users/Baiochi/Dropbox/USP/Lab/RawData/HCT/Targets.txt", header=TRUE, sep="\t", stringsAsFactors=FALSE)
raw = read.maimages(targets$FileName, source="agilent", green.only=TRUE, path="/Users/Baiochi/Dropbox/USP/Lab/RawData/HCT")
colnames(raw$E) <- targets$Nomeclature
colnames(raw$Eb) <- targets$Nomeclature
rownames(raw$targets) <- targets$Nomeclature
#setup printer
raw$genes$Block <- 1
names(raw$genes)[2] <- "Column"
raw$printer <- getLayout(raw$genes)
r <- raw$genes$Row
c <- raw$genes$Col
nr <- max(r)
nc <- max(c)
y <- rep(NA,nr*nc)
j <- (r-1)*nc+c
#y[j] <- log2(raw$E[,1])
#imageplot(y,raw$printer)
# Normal analysis
bgc = backgroundCorrect(raw,method='normexp')
norm = normalizeBetweenArrays(bgc,method='quantile')
eset = norm[norm$genes$ControlType==0,]
eset = avereps(eset,ID=eset$genes[,"SystematicName"])
f <- factor(targets$Condition, levels = unique(targets$Condition))
design = model.matrix(~0 + f)
colnames(design) <- levels(f)
fit <- lmFit(eset$E, design)
contrast.matrix = makeContrasts(contrasts='PM101-Ctr', levels=design)
fit <- contrasts.fit(fit, contrast.matrix)
fit <- eBayes(fit)
rank <- decideTests(fit,method="separate",adjust.method="fdr",p.value=0.05,lfc=0)
exprs = topTable(fit, adjust="fdr", coef='PM101-Ctr', genelist=eset$genes, number=Inf)
# Filtered Analysis
rawf <-raw
rawf$Eb[which(raw$Eb[,3]>60),3] <- round(mean(raw$Eb[,3]))
rawf$Eb[which(raw$Eb[,6]>60),6] <- round(mean(raw$Eb[,6]))
#rawf <- raw[ which(!(raw$Eb[,3]>60 | raw$Eb[,6]>60)) ,]
bgc2 = backgroundCorrect(rawf,method='normexp')
norm2 = normalizeBetweenArrays(bgc2,method='quantile')
eset2 = norm2[norm2$genes$ControlType==0,]
eset2 = avereps(eset2,ID=eset2$genes[,"SystematicName"])
fit2 <- lmFit(eset2$E, design)
contrast.matrix = makeContrasts(contrasts='PM101-Ctr', levels=design)
fit2 <- contrasts.fit(fit2, contrast.matrix)
fit2 = eBayes(fit2)
rank2 = decideTests(fit2,method="separate",adjust.method="fdr",p.value=0.05,lfc=0)
exprs2 = topTable(fit2, adjust="fdr", coef='PM101-Ctr', genelist=eset2$genes, number=Inf)
```

#Introduction

Background intensities from array 3 and 6 (labeled `PM101_3` and `Ctr_3`), which is from the same cell culture, have some probes with high values if compared to the other arrays. This can be considered an artifact due some random techinical problem.

In order to check the impact of this possible artifact, two types of analysis will be made and compared:
  
* Using all the probes  
* Filtering possible artifact spots. Each of these probes which background intesities are very higher will be replaced by the mean of their respective array


```{r}
kable(summary(raw$Eb), align='l', caption='Summary of the raw background signal')
```

***  

#Arrays visualization

  To see the image intensities from the arrays, values will be log2 transformed to a better visualization.

```{r Image Plot, fig.cap='Array images'}
par(mfrow=c(2,3))
for(i in 1:6){
  y[j] <- log2(raw$Eb[,i])
  imageplot(y, raw$printer)
}
```

  The very britgher spots on arrays 3 and 6 hide the true signal, we can see the difference if these probes are filtered.

```{r Image Plot2, fig.cap='Array images - filtered probes'}
par(mfrow=c(2,3))
for(i in 1:6){
  y[j] <- log2(rawf$Eb[,i])
  imageplot(y, raw$printer)
}
```

***  
  The max intensity value excluding these two arrays is 60, so we can see how many probes are highly expressed.

```{r , caption="Array 3"}
array3 <- raw[which(raw$Eb[,3]>60),]
table <- as.data.frame(cbind(array3$Eb[,3], array3$genes$GeneName,array3$genes$SystematicName))
colnames(table) <- c('Intensity', 'GeneName','SystematicName')
rownames(table) <- NULL
kable(table, caption='Array 3')
```
  
**Array PM101_3** have `r dim(array3)[1]` possible outlier probes
***  
```{r , caption='Array 6'}
array6 <- raw[which(raw$Eb[,6]>60),]
table <- as.data.frame(cbind(array6$Eb[,6], array6$genes$GeneName,array6$genes$SystematicName))
colnames(table) <- c('Intensity', 'GeneName','SystematicName')
rownames(table) <- NULL
kable(table, caption='Array 6')
```
  
**Array Ctr_3** have `r dim(array6)[1]` possible outlier probes
***  
And then see the p.values from these genes(in the not filtered analysis)
```{r}
out <- raw[ which(raw$Eb[,3]>60 | raw$Eb[,6]>60) ,]
artifact <- exprs[which(exprs$SystematicName %in% out$genes$SystematicName),]
rownames(artifact) <- NULL
kable(artifact[,c(9, 8, 16)]) #columns: SystematicName, GeneName and adj.P.Value
```
  
  As we can see, the only genes which has been ranked differentialy expressed is **NUMBL** and **BMPR1A** (adj.p.value < 0.05).
 
***  

#Analysis pipeline
  The following methods were used to perform both of the analysis:

**Background correction** using the method `normexp`:
```{r, eval=FALSE, results='hide', warning=FALSE}
bgc <- backgroundCorrect(raw,method='normexp')
```

**Normalization between the arrays** using the `quantile` method:
```{r, eval=FALSE, results='hide', warning=FALSE}
norm <- normalizeBetweenArrays(bgc,method='quantile')
```
**Filtering control probes**:
```{r, eval=FALSE, results='hide', warning=FALSE}
eset <- norm[norm$genes$ControlType==0,]
```
**Averagin replicated probes**:
```{r, eval=FALSE, results='hide', warning=FALSE}
eset <- avereps(eset,ID=eset$genes[,"SystematicName"])
```
**Create the linear model**:
```{r, eval=FALSE, results='hide', warning=FALSE}
f <- factor(targets$Condition, levels = unique(targets$Condition))
design <- model.matrix(~0 + f)
colnames(design) <- levels(f)
contrast.matrix <- makeContrasts(contrasts='PM101-Ctr', levels=design)
fit <- lmFit(eset$E, design)
```
**Compute empirical bayes statistics**:
```{r, eval=FALSE, results='hide', warning=FALSE}
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```

***  

### Foreground vs Background Plot
**All Probes**    
```{r, echo=FALSE, fig.width=9}
par(mfrow=c(2,3))
for(i in 1:6)
  plotFB(raw, array=i)
```
  
**Filtered Probes**   
```{r, echo=FALSE, fig.width=9}
par(mfrow=c(2,3))
for(i in 1:6)
  plotFB(rawf, array=i)
```

***  

### Raw Data Foreground Boxplot

Boxplot of the foregrounds intensities have no change
```{r, echo=FALSE, fig.width=9, fig.align='center'}
boxplot(log2(raw$E), main='Foreground Boxplot', col='firebrick2')
```

### Raw Data Background Boxplot

```{r, echo=FALSE, fig.width=9}
par(mfrow=c(1,2))
boxplot(log2(raw$Eb), main='Background Boxplot - All Probes', col='deepskyblue3')
boxplot(log2(rawf$Eb), main='Background Boxplot - Filtered Probes', col='forestgreen')
```

### Background Corrected Boxplot

```{r, echo=FALSE, fig.width=9}
par(mfrow=c(1,2))
boxplot(log2(bgc$E), main='Background Corrected Boxplot - All Probes', col='deepskyblue3')
boxplot(log2(bgc2$E), main='Background Corrected Boxplot - Filtered Probes', col='forestgreen')
```

### Post-Normalized Boxplot
The normalized boxplot is quite different between the analysis, the second diminishes the number of outliers values.
```{r, echo=FALSE, fig.width=9}
par(mfrow=c(1,2))
boxplot(norm$E, main='Normalized Boxplot - All Probes', col='deepskyblue3')
boxplot(norm2$E, main='Normalized Boxplot - Filtered Probes', col='forestgreen')
```

***  

### Densities Plots
```{r, echo=FALSE, warning=FALSE, fig.width=10}
par(mfrow=c(1,3))
plotDensities(raw, legend = 'topright', main='Raw Foreground densities - All Probes')
plotDensities(bgc, legend = 'topright', main='Raw Background densities - All Probes')
plotDensities(eset, legend = 'topright', main='Post-Normalized densities - All Probes')
```
```{r, echo=FALSE, warning=FALSE, fig.width=10}
par(mfrow=c(1,3))
plotDensities(rawf, legend = 'topright', main='Raw Foreground densities - Filtered Probes')
plotDensities(bgc2, legend = 'topright', main='Raw Background densities - Filtered Probes')
plotDensities(eset2, legend = 'topright', main='Post-Normalized densities - Filtered Probes')
```

***  

# Differential Expression

We can see the difference between the two analysis.
**All Probes**
Up regulated: `r as.numeric(table(rank)[3])`
Down regulated: `r as.numeric(table(rank)[1])`

**Filtered Probes**
Up regulated: `r as.numeric(table(rank2)[3])`
Down regulated: `r as.numeric(table(rank2)[1])`
  
###Volcanoplots


```{r, eval=FALSE}
par(mfrow=c(1,2))
volcanoplot(exprs$logFC, exprs$adj.P.Val, rank, title='PM101-Ctr All Probes')
volcanoplot(exprs2$logFC, exprs2$adj.P.Val, rank2, title='PM101-Ctr Filtered Probes')
```

```{r, echo=FALSE, fig.align='center', fig.height=9, fig.width=9}
volcanoplot(exprs$logFC, exprs$adj.P.Val, rank, title='PM101-Ctr All Probes')
```
```{r, echo=FALSE, fig.align='center', fig.height=9, fig.width=9}
volcanoplot(exprs2$logFC, exprs2$adj.P.Val, rank2, title='PM101-Ctr Filtered Probes')
```

diff <- exprs[exprs$adj.P.Val<0.05,]
diff2 <- exprs2[exprs2$adj.P.Val<0.05,]

From the design experiment with **all probes**, 
table(diff2$ProbeName %in% diff$ProbeName)




